---
title: Normalization of signalling proteins to total protein content
author: Arianne Richard and Claire Ma, with some code adapted from Aaron Lun
date: "`r Sys.Date()`"
output: html_document
---

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, warning=FALSE, message=FALSE)
```

# Overview

This script will build upon basic processing of the altered peptide ligand affinity mass cytometry data by normalizing each phosphorylated protein intensity to the intensity of its total protein measurement in each cell. To an artibtrary scaling factor (that would differ for each signalling molecule), this is analogous to finding the proportion of the total protein that is phosphorylated in each cell.

We start from a filtered, pre-gated dataset of fcs files.  Uncomment the line below and load the appropriate dataset.


# Loading the data

First we load the libraries and the data, narrowing channels to those with named markers for ease. 

```{r}
library(ncdfFlow)
library(cydar)
#fnames <- list.files(path=<directory with .fcs files>, pattern='.fcs', full.names=TRUE)

processed<-read.ncdfFlowSet(files = fnames)

channels <- data.frame(channel=colnames(processed), 
    marker=parameters(processed[[1]])$desc, 
    stringsAsFactors=FALSE)
channels
in.use <- grepl("^[0-9]+[A-Za-z]+_", channels$marker)
channels[in.use,]
processed <- processed[,in.use]

colnames(processed) <- as.character(parameters(processed[[1]])$desc)
```

# Further filtering

Normalization to total protein levels will only work in cells where total protein is detected.  We will perform a bit of transformation and further filtering on the processed data to exclude cells with low or missing levels of total protein detected. This requires a dataframe of phospho-protein channel names and their corresponding total protein channel names. Uncomment the appropriate line to create this dataframe, which will also be used later. 

Of note, the 3MAD lower cut-off for total protein expression may need to be altered depending on the distributions of total proteins in a given dataset.

```{r}

#paired_markers <- data.frame(phospho=c(<vector of phospho-protein markers>), total=c(<vector of total protein markers corresponding 1:1 with the phospho-protein markers>))
paired_markers

## transform the data

lgcl<-logicleTransform(w=0.1,t=10000) 
trans<-transformList(colnames(processed),lgcl)
processed <- transform(processed, trans)

## pool the cells

pool.ff <- poolCells(processed, equalize=FALSE)

## look at each total protein marker and make line at lower MAD 3 cut-off
gates <- vector('list', length=nrow(paired_markers))

for(i in 1:nrow(paired_markers)){
  hist(exprs(pool.ff)[,as.character(paired_markers$total[i])], breaks=200, 
       xlab=as.character(paired_markers$total[i]), main='')
  gates[i] <- outlierGate(pool.ff, as.character(paired_markers$total[i]), nmads=3, type='lower')
  abline(v=gates[[i]]@min, col='red')
}

# looks good - filter
for(i in 1:nrow(paired_markers)){
  processed <- Subset(processed, gates[[i]])
  pool.ff <- Subset(pool.ff, gates[[i]])
}

```

# Normalizing phospho- to total protein detected

Now we will see what happens when we normalize the phosphorylated protein intensites on a per cell basis to the total protein intensities.

First we need to revert to the raw intensity values and write the newly filtered files to disk.

```{r}
invlgcl <- inverseLogicleTransform(trans=lgcl)
trans<-transformList(colnames(processed),invlgcl)

processed2 <- transform(processed, trans)

fn <- rownames(processed2@phenoData)

for(i in 1:length(processed2)){
  write.FCS(processed2[[i]], filename=paste0('PROTfilt_', fn[i]))
}


```

Then we want to compute a ratio between the phosphorylated protein and the total protein levels. To maintain the scale similar to that measured by the antibodies, we'll multiply each phospho/total ratio by the median total protein expression measured in a selected sample. We'll replace the phosphorylated protein measurement with this ratio. The function below accomplishes this, taking as arguments the filtered, untransformed ncdfFlowSet, the phospho- and total protein dataframe, and the sample against which results will be scaled.

Uncomment the line running the function, enter parameters, and run.

```{r}

PROTnorm <- function(flowset, paired_markers, scale_sample){
  ## flowset = ncdfFlowSet object that has been filtered for cells in which total proteins are detectable
  ## paired_markers = dataframe of paired phospho and total protein markers (colnames are "phospho" and "total")
  ## scale_sample = name of sample that will be used to scale output ratios 
  ### so visually on same scale as normal antibody-measured intensities
  scalar <- flowset[[scale_sample]]
  for(i in 1:length(flowset)){
    ff <- flowset[[i]]
    for(j in 1:nrow(paired_markers)){
      pho <- exprs(ff)[,as.character(paired_markers$phospho[j])]
      tot <- exprs(ff)[,as.character(paired_markers$total[j])]
      scalar_tot <- exprs(scalar)[,as.character(paired_markers$total[j])]
      rat <- (pho/tot) * median(scalar_tot)
      exprs(ff)[,as.character(paired_markers$phospho[j])] <- rat
    }
    flowset[[i]] <- ff
  }
  return(flowset)
}

# processed2 <- PROTnorm(processed2, paired_markers, <scale_sample>)

```

Write the data for future use.

```{r}

for(i in 1:length(processed2)){
  write.FCS(processed2[[i]], filename=paste0('PROTnorm_', fn[i]))
}

```

# Wrapping up

```{r}
sessionInfo()
```
