---
title: Plots for 6-hour timecourse mass cytometry data after differential abundance analysis
author: Arianne Richard and Claire Ma
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Visualizing the data

First we load the data.

```{r}
library(cydar)
load('data/cydar_results_singlet_activation.RData')
```

## Creating dimensionality reduction plots

We'll create a tSNE plot of the DA hyperspheres.

```{r}
set.seed(100)
library(Rtsne)
t.out <- Rtsne(intensities(cd2), perplexity=30) 
head(t.out$Y)
```

We can then colour the coordinates by the desired log-fold-changes. 

## ANOVA-like comparison

Only the hyperspheres with significant changes at any time points are shown here. We've maxed out the logFC at 6 in order to be able to see the subtle changes since many of the changes are VERY large, going from no activated cells to mostly activated cells.

```{r}
png('data/DA_tSNE.png',height=9, width = 10, units="in", res=200)
layout(cbind(matrix(c(1:16), nrow=4),17), widths=c(4, 4, 4, 4, 1))
par(mar=c(2.1, 2.1, 2.1, 1.1), oma=c(3,3,1,1), mgp=c(2.2,1,0))
pt <- c('NP68_1', 'G4_1', 'T4_1', 'N4_1', 'NP68_2', 'G4_2', 'T4_2', 'N4_2', 'NP68_4', 'G4_4', 'T4_4', 'N4_4', 'NP68_6', 'G4_6', 'T4_6', 'N4_6')

for(peptide_time in pt){
    fc.col <- plotSphereLogFC(t.out$Y[is.sig.all,1], t.out$Y[is.sig.all,2], 
	    logFC=res.all$table[is.sig.all,paste0("logFC.peptide_time",peptide_time)],
        max.logFC=6, cex.axis=1.5, cex.lab=1.5, cex.main=1.5, cex=1, main=paste0(peptide_time, 'h'))
}
par(mar=c(0,0,0,0))
plot(0,0, axes=FALSE, xlab="", ylab="")
createColorBar(fc.col)
dev.off()
```

To annotate them, we will make similar plots coloured by the intensity of all relevant markers.

```{r}
png("data/DA_tSNE_intensities.png",height=9, width = 10, units="in", res=200)
par(oma=c(4.1,4.1,1.1,1.1))
lmat <- cbind(matrix(seq_len(4*4), ncol=4), 17)
layout(lmat, widths=c(rep(4, 4), 1))

reranges <- intensityRanges(cd2)
par(mar=c(2.1, 2.1, 2.1, 2.1))

for (i in order(markernames(cd2))) {
    int.col <- plotSphereIntensity(t.out$Y[is.sig.all,1], t.out$Y[is.sig.all,2], 
        intensities(cd2)[is.sig.all,i], irange=reranges[,i], 
        main=markernames(cd2)[i], 
        cex=1, cex.lab=1.5, cex.main=1.5, cex.axis=1.5)
}

# just filling up the leftover panels. 
for (j in seq_len(16-ncol(intensities(cd2)))) { plot.new() } 

par(mar=c(0,0,0,0))
plot(0,0, type="n", axes=FALSE, ylab="", xlab="")
createColorBar(int.col, top.text="High", bottom.text="Low")
dev.off()
```

An alternative visualization strategy is to create heatmaps for each hypersphere.
First we add some row names to our intensity matrix (otherwise `pheatmap()` becomes quite unhappy).

```{r}
library(pheatmap)
mat <- intensities(cd2)
rownames(mat) <- seq_len(nrow(mat)) 
```

We indicate whether each hypersphere was significant, and if so the direction in which its abundance changed for each condition versus unstimulated cells.

```{r}
ptr <- rev(pt)

statuses <- list()
for (i in 1:length(ptr)) {
  coln <- paste0("logFC.peptide_time",ptr[i])
  cur.lfc <- res.all$table[,coln]
  cur.sign <- rep('small_change', length(cur.lfc))
  cur.sign[cur.lfc > 6] <- 'more_than_6'
  cur.sign[(cur.lfc <= 6) & (cur.lfc > 4)] <- 'more_than_4'
  cur.sign[(cur.lfc <= 4) & (cur.lfc > 2)] <- 'more_than_2'
  cur.sign[cur.lfc < -6] <- 'less_than_n6'
  cur.sign[(cur.lfc >= -6) & (cur.lfc < -4)] <- 'less_than_n4'
  cur.sign[(cur.lfc >= -4) & (cur.lfc < -2)] <- 'less_than_n2'
  cur.sign[!is.sig.all] <- "ns"
  statuses[[i]] <- cur.sign
}

names(statuses) <- ptr

library(RColorBrewer)
cols <- colorRampPalette(c("red", "pink", "lightgray","lightblue", "blue"))(7)

# Adding colors for each status.
fc.colors <- c(more_than_6=cols[1], more_than_4=cols[2], more_than_2=cols[3], small_change=cols[4], less_than_n2=cols[5], less_than_n4=cols[6], less_than_n6=cols[7], ns='gray30')
fc.colors <- rep(list(fc.colors), length(statuses))
names(fc.colors) <- names(statuses)    
```

Finally, we create a heatmap using pheatmap and clustering the marker intensities within the hyperspheres by Pearson corrleation or Euclidean distance.

```{r}
set.seed(100)
pdf("data/DA_heatmap_correlation.pdf", height=10, width=7)
pheatmap(mat,
         clustering_distance_rows="correlation",
         clustering_distance_columns="correlation",
         col=viridis::viridis(101),
         cluster_cols=TRUE,
         annotation_row=do.call(data.frame, statuses),
         annotation_color=fc.colors,
         show_rownames=FALSE)
dev.off()

set.seed(100)
pdf("data/DA_heatmap_euclidean.pdf", height=10, width=7)
pheatmap(mat,
         clustering_distance_rows="euclidean",
         clustering_distance_columns="euclidean",
         col=viridis::viridis(101),
         cluster_cols=TRUE,
         annotation_row=do.call(data.frame, statuses),
         annotation_color=fc.colors,
         show_rownames=FALSE)
dev.off()
```

# Wrapping up

```{r}
sessionInfo()
```
